Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2014-11-07 draftfinalreport.Rnw
hmmm

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{rotating}
\usepackage{threeparttable}
\usepackage{blindtext}
\usepackage{dcolumn}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
%\usepackage{caption2}
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{placeins}
\usepackage{framed}
\usepackage{float}
\usepackage{authblk}
\newcommand{\vecA}{\mathbf{A}}
\newcommand{\vecB}{\mathbf{B}}
\newcommand{\vecC}{\mathbf{C}}
\newcommand{\vecD}{\mathbf{D}}
\newcommand{\vecE}{\mathbf{E}}
\newcommand{\vecF}{\mathbf{F}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecI}{\mathbf{I}}
\newcommand{\vecJ}{\mathbf{J}}
\newcommand{\vecK}{\mathbf{K}}
\newcommand{\vecM}{\mathbf{M}}
\newcommand{\vecN}{\mathbf{N}}
\newcommand{\vecP}{\mathbf{P}}
\newcommand{\vecQ}{\mathbf{Q}}
\newcommand{\vecS}{\mathbf{S}}
\newcommand{\vecT}{\mathbf{T}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecZ}{\mathbf{Z}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecc}{\mathbf{c}}
\newcommand{\vece}{\mathbf{e}}
\newcommand{\vecf}{\mathbf{f}}
\newcommand{\vecg}{\mathbf{g}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\veci}{\mathbf{i}}
\newcommand{\vecr}{\mathbf{r}}
\newcommand{\vecu}{\mathbf{u}}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecw}{\mathbf{w}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vecz}{\mathbf{z}}
\newcommand{\Ach}{\mathcal{H}}
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\En}{\mathcal{N}}
\newcommand{\Pe}{\mathcal{P}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Rc}{\mathcal{R}_c}
\newcommand{\Pv}{\mathcal{P}_v}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Qu}{\mathcal{Q}}
\newcommand{\Te}{\mathcal{T}}
\newcommand{\Ve}{\mathcal{V}}
\newcommand{\Dub}{\mathcal{W}}
\newcommand{\Zed}{\mathcal{Z}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\de}{\mbox{d}}
\newcommand{\wrt}{\,\mbox{d}}
\newcommand{\veczero}{\mathbf{0}}
\newcommand{\vecone}{\mathbf{1}}
\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
\newcommand{\cosech}{\mbox{cosech}\,}
\newcommand{\sign}{\mbox{sign}\,}
\newcommand{\trace}{\mbox{trace}\,}
\newcommand{\Inf}{\mbox{Inf}\,}
%\newcommand{\Ro}{R_0}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Those under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.

We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{section:epidemic_modelling}).

Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\ref{section:epidemic_modelling}). We estimate the costs of previous and the current measles outbreaks (section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\ref{sub:cost_benefit}). This report includes further developments of the analyses provided in previous interim reports.

@@

We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}). In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand. To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{{sub:imp_risk}}).

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).

@@

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 15,height=11>>=
par(mar=c(12,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="",cex.lab=2,xaxt="n")
axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
mtext(side=1,line=10,text="Age and NZDep",cex=2)
legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
par(mar = c(5,4,4,2) + 0.1)
@
\end{center}
\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

@@

In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR. We describe the population immunity levels at both a national level and for each DHB in New Zealand. To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work. Specifically, we use the data in Table \ref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
@
\end{center}
\caption{Distribution of measles cases per category (Table~\ref{table:percap}) used in the final regression model (Equation~\ref{eq:reg})}
\label{fig:histcasecat}
\end{figure}

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
@
\end{center}
\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the case data (Table~\ref{table:percap})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
\label{fig:resid}
\end{figure}

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
library(xtable)
resan<-xtable(anova(model3,test="F"))
print(resan,floating = FALSE)
@
\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space

A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

@@

\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\ref{fig:ageinyears}).

The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and section~\ref{sub:popim}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\ref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\ref{fig:vaccstat}).

@@

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
plot(pop,xlab="Age",ylab="Population",cex.lab=2)
points(naive,pch=16)
legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
@
\end{center}
\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\ref{table:sero})}
\label{fig:naive}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
     \label{fig:naive_map}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
     \label{fig:Northland}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
     \label{fig:Waitemata}
\end{figure}

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage{subfig}
\usepackage{enumitem}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{rotating}
\usepackage{threeparttable}
\usepackage{blindtext}
\usepackage{dcolumn}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{caption2}
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{placeins}
\usepackage{framed}
\usepackage{float}
\usepackage{authblk}
\newcommand{\vecA}{\mathbf{A}}
\newcommand{\vecB}{\mathbf{B}}
\newcommand{\vecC}{\mathbf{C}}
\newcommand{\vecD}{\mathbf{D}}
\newcommand{\vecE}{\mathbf{E}}
\newcommand{\vecF}{\mathbf{F}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecI}{\mathbf{I}}
\newcommand{\vecJ}{\mathbf{J}}
\newcommand{\vecK}{\mathbf{K}}
\newcommand{\vecM}{\mathbf{M}}
\newcommand{\vecN}{\mathbf{N}}
\newcommand{\vecP}{\mathbf{P}}
\newcommand{\vecQ}{\mathbf{Q}}
\newcommand{\vecS}{\mathbf{S}}
\newcommand{\vecT}{\mathbf{T}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecZ}{\mathbf{Z}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecc}{\mathbf{c}}
\newcommand{\vece}{\mathbf{e}}
\newcommand{\vecf}{\mathbf{f}}
\newcommand{\vecg}{\mathbf{g}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\veci}{\mathbf{i}}
\newcommand{\vecr}{\mathbf{r}}
\newcommand{\vecu}{\mathbf{u}}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecw}{\mathbf{w}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vecz}{\mathbf{z}}
\newcommand{\Ach}{\mathcal{H}}
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\En}{\mathcal{N}}
\newcommand{\Pe}{\mathcal{P}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Rc}{\mathcal{R}_c}
\newcommand{\Pv}{\mathcal{P}_v}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Qu}{\mathcal{Q}}
\newcommand{\Te}{\mathcal{T}}
\newcommand{\Ve}{\mathcal{V}}
\newcommand{\Dub}{\mathcal{W}}
\newcommand{\Zed}{\mathcal{Z}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\de}{\mbox{d}}
\newcommand{\wrt}{\,\mbox{d}}
\newcommand{\veczero}{\mathbf{0}}
\newcommand{\vecone}{\mathbf{1}}
\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
\newcommand{\cosech}{\mbox{cosech}\,}
\newcommand{\sign}{\mbox{sign}\,}
\newcommand{\trace}{\mbox{trace}\,}
\newcommand{\Inf}{\mbox{Inf}\,}
%\newcommand{\Ro}{R_0}
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Those under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\autoref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.

We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our modelling and risk analyses results (section~\autoref{sub:risk_analyses} and section~\autoref{section:epidemic_modelling}).

Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\autoref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\autoref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\autoref{section:epidemic_modelling}). We estimate the costs of previous and the current measles outbreaks (section~\autoref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\autoref{sub:cost_benefit}). This report includes further developments of the analyses provided in previous interim reports.

@@

We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\autoref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\autoref{sub:immunestat}), and current rates of immunity within the population (section~\autoref{sub:immunestat}). In light of the apparent increasing trend in measles incidence in the last few years (Figure~\autoref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand. To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\autoref{{sub:imp_risk}}).

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\autoref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).

@@

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\autoref{table:none}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\autoref{fig:percap}, Table~\autoref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\autoref{table:percap}).

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 15,height=11>>=
par(mar=c(12,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="",cex.lab=2,xaxt="n")
axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
mtext(side=1,line=10,text="Age and NZDep",cex=2)
legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
par(mar = c(5,4,4,2) + 0.1)
@
\end{center}
\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\autoref{table:percap} for details)}
\label{fig:percap}
\end{figure}

@@

In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR. We describe the population immunity levels at both a national level and for each DHB in New Zealand. To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work. Specifically, we use the data in Table \autoref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\autoref{fig:histcasecat}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
@
\end{center}
\caption{Distribution of measles cases per category (Table~\autoref{table:percap}) used in the final regression model (Equation~\autoref{eq:reg})}
\label{fig:histcasecat}
\end{figure}

The predicted values from the regression model plotted against the reported cases are shown in Figure~\autoref{fig:predict}, and the residuals are shown in Figure~\autoref{fig:resid}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
@
\end{center}
\caption{Regression model (Equation~\autoref{eq:reg}) predictions plotted against the case data (Table~\autoref{table:percap})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\autoref{eq:reg})}
\label{fig:resid}
\end{figure}

The significance of the different predictor variables can be seen in the ANOVA results (Table~\autoref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
library(xtable)
resan<-xtable(anova(model3,test="F"))
print(resan,floating = FALSE)
@
\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\autoref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space

A summary of the regression model (Equation~\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\autoref{table:regression}.

@@

\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\autoref{fig:ageinyears}).

The majority of measles cases are among people of European ancestry (Table~\autoref{table:percap} and Figure~\autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\autoref{fig:ageinyears} and ~\autoref{fig:ageandvac}, Table~\autoref{table:sero}, and section~\autoref{sub:popim}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\autoref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\autoref{fig:ageandvac}). However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second. A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\autoref{fig:vaccstat}).

@@

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
plot(pop,xlab="Age",ylab="Population",cex.lab=2)
points(naive,pch=16)
legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
@
\end{center}
\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\autoref{table:sero})}
\label{fig:naive}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\autoref{table:sero})}
     \label{fig:naive_map}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\autoref{table:sero})}
     \label{fig:Northland}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\autoref{table:sero})}
     \label{fig:Waitemata}
\end{figure}

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
     \end{center}
     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\autoref{table:sero})}
     \label{fig:Auckland}
\end{figure}
